---
title: "ARIMA Analysis of Monthly Average Retail Electricity Prices"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Description

The purpose of this projects is to explore the use of Autoregressive Integrated Moving Average (ARIMA) models to forecast the monthly average electricity prices for the Pacific Contiguous Region.

## Load the tidyverse and forecast packages

```{r, message=FALSE}
library(tidyverse)
library(forecast)
```

## Read in the monthly average electricity data for the Pacific Contiguous Region (cents per kilawatthour)

The data was downloaded from the [U.S. Energy Information Administration website](https://www.eia.gov/electricity/data/browser/#/topic/7?agg=0,1&geo=01&endsec=vg&linechart=ELEC.PRICE.PCC-ALL.M&columnchart=ELEC.PRICE.PCC-ALL.M&map=ELEC.PRICE.PCC-ALL.M&freq=M&ctype=linechart&ltype=pin&rtype=s&pin=&rse=0&maptype=0`) for the time period of Jan-2001 through Oct-2020.  However, only data through Dec-2019 was used since the 2020 data was not considered final at the time of this analysis.

```{r, message=FALSE}
series <- read_csv("https://raw.githubusercontent.com/tbstegman/ARIMA_Analysis_of_Electricity_Prices/main/Average_retail_price_of_electricity_monthly.csv",skip=4)
```

## Perform some basic transformations and plot the data

```{r tsplot, message=FALSE}
# Rename the variables
names(series) <- c("month","avg_price")

# Convert month to Date class
series$month <- as.Date.character(paste("01",series$month),format="%d %b %Y")

# Sort the data by ascending month
series <- series[order(series$month),]

# Create a time series object of the monthly data but exclude the last 12 months
price <- ts(series$avg_price,frequency=12,start=c(2001,1)) %>% window(end=c(2019,12))

# Plot the time series
autoplot(price) +
  labs(x="Year",y="Avg. Price (cents per kilawatthour)") +
  ggtitle("Average Electricity Prices for the Pacific Contiguous Region") +
  theme(plot.title = element_text(hjust = 0.5,size=15))
```

From this plot it looks like the variance is fairly stable.  However, we can see both a pronounced upward trend as well as a seasonal pattern, which indicate that the series is not stationary.

## Create a training data set, and hold out the last 24 months of data for testing

```{r}
train <- window(price, start=c(2001,1), end=c(2017,12))
test <- window(price, start=c(2018,1))
```

Now, run the **ndiffs** and **nsdiffs** functions on the training set to determine the number of differences suggested to stabilize the mean and make the series stationary.

```{r}
ndiffs(train)
nsdiffs(diff(train))
```

Based upon the output from the **ndiffs** and **nsdiffs** functions, it looks like a first difference along with a difference at lag 12 would be needed to make the series stationary.  So, we apply these differences and produce the time series, ACF, and PACF plots for the differenced series.

```{r train_diff_plt}
# Differnce the series
diff1 <- diff(train)
diff1_12 <- diff(diff1,lag=12)

# Create plots for the differenced data
ggtsdisplay(diff1_12)
```

The time series plot looks more or less stationary.  Since the ACF drops off sharply at lag 1, and the PACF decays more slowly, it looks as though a non-seasonal MA(1) term might be indicated.  Also, since the ACF has a pronounced spike at lag 12, but is below the threshold at lags 24 and 36, while the PACF drops off more slowly at these seasonal lags, a seasonal MA(1) might also be appropriate.  Putting these pieces of information together, it appears that an ARIMA(0,1,1)(0,1,1)<sub>12</sub> model might be a good fit for this series.

## Use the auto.arima function to find the best models for the AIC, AICc, and BIC information criterion

Use auto.arima to run through all possible orders with d=1 and D=1 up to ARIMA(4,1,4)(2,1,2)<sub>12</sub> to find the best models based upon the AICc, AIC, and BIC information criterion.

``` {r}
# Find the model that minimizes AICc
Arimafit1 <- auto.arima(train, d=1, D=1, ic="aicc", max.p=4, max.q=4, max.order=14, stepwise=F, approximation=F)
summary(Arimafit1)

# Find the model that minimizes AIC
Arimafit2 <- auto.arima(train, d=1, D=1, ic="aic", max.p=4, max.q=4, max.order=14, stepwise=F, approximation=F)
summary(Arimafit2)

# Find the model that minimizes BIC
Arimafit3 <- auto.arima(train, d=1, D=1, ic="bic", max.p=4, max.q=4, max.order=14, stepwise=F, approximation=F)
summary(Arimafit3)
```

The model which minimized the AICc and AIC was found to be an ARIMA(1,1,1)(0,1,1)<sub>12</sub> model, while the model that minimized the BIC was ARIMA(0,1,1)(0,1,1)<sub>12</sub> (which coincides with the model identified earlier based upon the ACF and PACF patterns).

## Create a function to check residuals, and calculate forecast accuracy metrics on the test data for a given ARIMA model fit
```{r}
ARIMA_checks <- function(fit){
  checkresiduals(fit)
  accuracy(forecast(fit,h=length(test)),test)
}
```

Perform checks for the two ARIMA models identified above.
```{r ARIMA1_checks}
ARIMA_checks(Arimafit1)
```

```{r ARIMA3_checks}
ARIMA_checks(Arimafit3)
```

Both models fit the training data pretty well.  The residuals for both series appear to be white noise, and the the forecast accuracy metrics are comparable for the two models.

Now, perform the same checks for the benchmark model of ARIMA(0,1,0)(0,1,0)<sub>12</sub> to compare its performance against the two ARIMA models identified above.
```{r bench_checks}
bench <- Arima(train,order=c(0,1,0),seasonal=c(0,1,0))
ARIMA_checks(bench)
```

Based upon the diagnostic plots, Ljung-Box test, and forecast accuracy metrics, we can see that the two ARIMA models identified earlier both fit the data better than the benchmark model.

## Choose the best model, and produce a forecast with prediction intervals for the next 12 months

Since both the ARIMA(1,1,1)(0,1,1)<sub>12</sub> and ARIMA(0,1,1)(0,1,1)<sub>12</sub> models have excellent diagnostics and perform similarly well on the test set, we choose the ARIMA(0,1,1)(0,1,1)<sub>12</sub> model for forecasting as it is the more parsimonious of the two.

```{r final}
# Fit the best model to all of the data
best <- Arima(price, order=c(0,1,1), seasonal=c(0,1,1))

# Print the forecast
forecast(best,h=12)

# Plot the forecast
autoplot(forecast(best,h=12)) +
  labs(x="Year",y="Avg. Price (cents per kilawatthour)") +
  theme(plot.title = element_text(hjust = 0.5,size=15))
```


## References

Hyndman, R.J., & Athanasopoulos, G. (2018) Forecasting: principles and practice, 2nd edition, OTexts: Melbourne, Australia. OTexts.com/fpp2. Accessed on 04Jan2021.

Hyndman R, Athanasopoulos G, Bergmeir C, Caceres G, Chhay L, O'Hara-Wild M, Petropoulos F,Razbash S, Wang E, Yasmeen F (2020). _forecast: Forecasting functions for time series and linear models_. R package version 8.13, <URL: https://pkg.robjhyndman.com/forecast/>.

Hyndman RJ, Khandakar Y (2008). “Automatic time series forecasting: the forecast package for R.” _Journal of Statistical Software_, *26*(3), 1-22. <URL: https://www.jstatsoft.org/article/view/v027i03>.
